AGB Stars in the IR: Physical Characteristics, Galactic Spatial Distribution, and Chemical Impact of the Milky Way AGB Population

### 2015-05-05 paper1/paper1_criteria.tex
Update to paper 1

\section{Object Selection Criteria}
\label{sec:criteria}
We create color-color criteria to generate a catalog of AGB candidates. We seek to maximize AGB completeness while minimizing contamination from non-AGB objects to beneath the 1\% level.

The color-color criteria for AGB selection are as follows:
\begin{align} 
(J-K_s) > 1.1\label{eq:criteria1}\\
(W2-W3) > 0.3\label{eq:criteria2}\\
(W3-W4) < -0.83(W2-W3) + 3.37\label{eq:criteria3}
\end{align}
The criteria in (\ref{eq:criteria1}) and (\ref{eq:criteria2}) are primarily concerned with rejecting objects from the stellar locus, and other stars whose NIR spectra are dominated by the Rayleigh-Jeans tail (many covered by the sample of galaxies). (\ref{eq:criteria3}) also rejects stars from the stellar locus, but primarily functions to remove IR-bright extragalactic sources.

\noindent Sample completeness $\eta$ is defined as
\begin{eqnarray*}
\eta &=& \frac{N - n_\text{missed}}{N}
\end{eqnarray*}
where $N$ is the total number of objects in the sample, and $n_\text{missed}$ is the number of objects outside of the applied boundaries. Figure~\ref{fig:completeness} shows the distribution of sample completeness amongst AGB sources after the application of the above criteria. The vast majority (79.07\%) of Galactic AGB stars is recovered after criteria (\ref{eq:criteria1}), (\ref{eq:criteria2}), and (\ref{eq:criteria3}) are applied. The largest losses occur at the edges of the Galactic disk ($\lvert b\rvert\approx10^\circ$) and in regions of high stellar number density both in the Galaxy and the Magellanic Clouds. In the color-color diagrams, selection completeness degrades near the borders of the selction area, as objects that straddle these boundaries may exhibit enough emission in other color-color spaces to be removed by our criteria. Of the remaining sample of 5,709 objects, 52.94\% lie within the disk ($\lvert b\rvert<10^\circ$).

@@

\cite{2013sdmm.book.....I} defines sample contamination as
\begin{eqnarray*}
\epsilon &=& \frac{n_\text{spurious}}{n_\text{selected}}.
\end{eqnarray*}
where $n_\text{spurious}$ is the number of spurious sources and $n_\text{selected} = N + n_\text{spurious}$. The contamination map is shown in Figure~\ref{fig:contamination}. Most bins in Figure~\ref{fig:contamination} exhibit 0\% contamination, with the overall contamination level at 0.38\%. What contaminants do remain exist primarily at the very fringes of the AGB star distribution in ($J-K_S$) vs ($W2 - W3$) space and near the redder boundary in ($W3-W4$) vs ($W2 - W3$), where bluer extragalactic sources creep into the selection region. The results of the applied critiera on both the collective AGB and contaminant samples are summarized in Table~\ref{tab:completeness}.

@@

>>>>

\section{Object Selection Criteria}
\label{sec:criteria}
In creating color-color criteria to generate a catalog of AGB candidates, we seek to maximize AGB completeness while minimizing contamination from non-AGB objects to beneath the 1\% level.

The color-color criteria for AGB selection are as follows:
\begin{align} 
	(J-K_s) > 1.1\label{eq:criteria1}\\
	(W2-W3) > 0.3\label{eq:criteria2}\\
	(W3-W4) < -0.83(W2-W3) + 3.37\label{eq:criteria3}
\end{align}
The criteria in (\ref{eq:criteria1}) and (\ref{eq:criteria2}) are primarily concerned with rejecting objects from the stellar locus, and other objects whose NIR spectra are dominated by the Rayleigh-Jeans tail. (\ref{eq:criteria3}) also rejects stars from the stellar locus, but primarily functions to remove IR-bright extragalactic sources.

We also experiment with other criteria to focus more closely on the high-reliability AGB population, instead of trying to capture the most AGB-like objects. These criteria are as follows:
\begin{align} 
	(W1-W2) < 1\label{eq:criteria4}\\
	(W2-W3) < 1\label{eq:criteria5}
\end{align}
Criteria (\ref{eq:critera4}) restricts NIR excesses. Criteria (\ref{eq:criteria5}) primarily removes extragalactic sources, while only slightly cutting into the distribution of known AGB stars.

\noindent Sample completeness $\eta$ is defined as
\begin{eqnarray*}
\eta &=& \frac{N - n_\text{missed}}{N}
\end{eqnarray*}
where $N$ is the total number of objects in the sample, and $n_\text{missed}$ is the number of objects outside of the applied boundaries. Figure~\ref{fig:completeness} shows the distribution of sample completeness amongst AGB sources after the application of the above criteria. The vast majority (79.07\%) of Galactic AGB stars is recovered after criteria (\ref{eq:criteria1}), (\ref{eq:criteria2}), and (\ref{eq:criteria3}) are applied. The largest losses occur at the edges of the Galactic disk ($\lvert b\rvert\approx10^\circ$) and in regions of high stellar number density both in the Galaxy and the Magellanic Clouds. In the color-color diagrams, selection completeness degrades near the borders of the selection area, as objects that straddle these boundaries may exhibit enough emission in other color-color spaces to be removed by our criteria. Of the remaining sample of 5,709 objects, 52.94\% lie within the disk ($\lvert b\rvert<10^\circ$).

\begin{table}[h]
    \begin{center}
        \caption{Samples Recovered wih Application of Criteria (\%)}
        \scalebox{0.85}{
            \begin{tabular}{l r r r r r r r r r r r r}
\hline
Object Type & Reduced & (1) & (2) & (3) & (1,2) & (1,2,3) & | & \textbf{(4)} & \textbf{(5)} & \textbf{(1,2,4)} & \textbf{(1,2,5)} & \textbf{(1,2,4,5)}\\
\hline
All AGBs & 7220 & 95.15 & 84.89 & 95.21 & 82.62 & 79.07 & | & \textbf{94.22} & \textbf{88.82} & \textbf{77.05} & \textbf{72.71} & \textbf{71.14}\\ 
O-rich AGB & 3147 & 98.98 & 100.00 & 100.00 & 98.98 & 98.98 & | & \textbf{98.44} & \textbf{97.33} & \textbf{97.43} & \textbf{96.31} & \textbf{94.88}\\ 
C-rich AGB & 540 & 99.26 & 99.81 & 99.26 & 99.07 & 98.33 & | & \textbf{57.59} & \textbf{63.33} & \textbf{56.85} & \textbf{62.59} & \textbf{54.44}\\ 
Unclassed AGB & 3533 & 91.11 & 69.15 & 90.32 & 65.53 & 58.39 & | & \textbf{96.07} & \textbf{85.14} & \textbf{61.99} & \textbf{53.24} & \textbf{52.53}\\ 
DR12 SSPP & 67508 & 76.16 & 99.16 & 0.91 & 76.07 & 0.05 & | & \textbf{86.05} & \textbf{0.89} & \textbf{65.67} & \textbf{0.03} & \textbf{0.03}\\ 
DR7 LRG & 84 & 96.43 & 100.00 & 0.00 & 96.43 & 0.00 & | & \textbf{89.29} & \textbf{0.00} & \textbf{85.71} & \textbf{0.00} & \textbf{0.00}\\ 
QSO & 3783 & 73.54 & 99.97 & 0.13 & 73.54 & 0.11 & | & \textbf{34.50} & \textbf{0.11} & \textbf{27.54} & \textbf{0.08} & \textbf{0.08}\\ 
Galaxy & 42066 & 78.55 & 100.00 & 0.01 & 78.54 & 0.00 & | & \textbf{96.39} & \textbf{0.01} & \textbf{75.31} & \textbf{0.00} & \textbf{0.00}\\ 
AGN & 1011 & 79.33 & 100.00 & 0.00 & 79.33 & 0.00 & | & \textbf{96.34} & \textbf{0.00} & \textbf{75.87} & \textbf{0.00} & \textbf{0.00}\\

\hline
                \label{tab:criteria}
            \end{tabular}}    
    \end{center}
\end{table}

@@

\cite{2013sdmm.book.....I} defines sample contamination as
\begin{eqnarray*}
\epsilon &=& \frac{n_\text{spurious}}{n_\text{selected}}.
\end{eqnarray*}
where $n_\text{spurious}$ is the number of spurious sources and $n_\text{selected} = N + n_\text{spurious}$. The contamination map is shown in Figure~\ref{fig:contamination}. Most bins in Figure~\ref{fig:contamination} exhibit 0\% contamination, with the overall contamination level at 0.38\%. What contaminants do remain exist primarily at the very fringes of the AGB star distribution in ($J-K_S$) vs ($W2 - W3$) space and near the redder boundary in ($W3-W4$) vs ($W2 - W3$), where bluer extragalactic sources creep into the selection region. The results of the applied criteria on both the collective AGB and contaminant samples are summarized in Table~\ref{tab:completeness}.

We note that while we do achieve a low rate of contamination, most of our contaminant sources are out of the plane of the Galaxy, while many of our \agb\, sources are within the Galactic disk. Considering how our extragalactic contaminant sources are from various iterations of \sdss, we know that their spatial distributions should be effectively uniform across the sky {\color{red} [cite me]}. For contaminants unaffected by interstellar reddening, they would occupy the same regions of color-color space already marked by our existing contaminant sample, and would similarly be removed by (\ref{eq:criteria1}), (\ref{eq:criteria2}), and (\ref{eq:criteria3}). This would produce the same zero-level contamination for all extragalactic sources aside from QSOs. We can adopt a QSO spatial density by dividing the total number of spectroscopically identified QSOs by the covered area on the sky. Using this, we recover a QSO spatial density of {\color{red} some number per square degree}. Using that density, and assuming ubiquity of sources on the sky, complete detection of QSOs {\color{red} down to some magnitude limit}, and a lack of interstellar reddening, we can estimate a maximum QSO contamination fraction of {\color{red} some number}. We use this number to report our QSO contamination fraction in Table~\ref{tab:completeness}, as QSOs significantly affected by interstellar reddening would exist further outside of the limits of (\ref{eq:criteria1}), (\ref{eq:criteria2}), and (\ref{eq:criteria3}). A more rigorous study of QSO spatial density is outside the scope of this paper.

The remaining contaminant sample is from the \sdss\, DR12 SSPP, with the number of remaining species found within shown in Figure~\ref{fig:sspp_histo}. {\color{red} the figures don't match up with each other. The number of red stars on the right side of the fig are significantly less than those in the histogram. Reconcile this.}

@@

\begin{figure}[h]
\includegraphics[width=6.7in]{figs/contaminating_sspp_objects.pdf}
\includegraphics[width=6.7in]{figs/contaminating_sspp_objects_classes.pdf}
\caption{Some histogram\label{fig:sspp_histo}}
\end{figure}

### 2015-05-05 paper1/paper1_intro.tex
Update to paper 1

\section{Introduction}
The formation of galaxies like the \mw\, was long thought to be a steady process that created a smooth distribution of stars. This standard view was exemplified by the \cite{1980ApJS...44...73B} and \cite{1989ARA&A..27..555G} models, described in detail by, e.g., \cite{1993ARA&A..31..575M}. It was further motivated by observations of other galaxies, as well as what little information was available from \emph{High Precision Parallax Collecting Satellite} \citep[\emph{HIPPARCOS}, ][]{1984SSRv...39....1K} and smaller pencil-beam surveys. In these, the \mw\, is modeled by three discrete components described by simple analytic expressions: the thin disk, thick disk, and halo.

The advent of the \emph{Sloan Digital Sky Survey} \citep[\sdss, ][]{2000AJ....120.1579Y} alleviated these limitations, providing accurate digital multi-band optical photometry across a quarter of the sky, as well as the largest optical spectroscopic catalog thus far known. This new influx of data enabled the development and application of photometric parallax methods, using color-magnitude relations to estimate stellar distances. In turn, this led to the large scale ``tomography" of the \mw\; via stellar distributions in the 7-dimensional space spanned by spatial coordinates \citep{2008ApJ...673..864J}, velocity components \citep{2010ApJ...716....1B}, and metallicity \citep{2008ApJ...684..287I}. The resulting maps revealed rich, complex substructure in the distribution of the \mw's stars \citep[e.g.][]{2000AJ....120..963I,2000ApJ...540..825Y,2001ApJ...554L..33V,2002ApJ...569..245N,2003ApJ...599.1082M,2006ApJ...642L.137B,2006ApJ...651L..29G,2006AJ....132..714V}, deeply shaking the existing view of the Galaxy.

In order to move forward from where \sdss\, tomography left off, we require observations that span an area larger than that of \sdss\, with Galactic objects that can be seen through interstellar dust out to large distances. Stars from the Asymptotic Giant Branch (\agb) are perfect candidates as probes to the \mw. \agb\, stars are represent the last stage of evolution for stars between 0.8 and 8 $M_\odot$ \citep{1983ARA&A..21..271I, 2005ARA&A..43..435H}, so they are bound to reside throughout the galaxy wherever other stars are present. During this phase, they produce substantial stellar winds \citep[$10^{-7} < \dot{M} < 10^{-4} M_\odot$ yr$^{-1}$,][]{2002A&A...391.1053O} rich in SiO and amorphous carbon as they progress through being oxygen-rich to being carbon-rich. These winds collect into circumstellar shells that, when warmed by the stellar photosphere, shine brightly in the near- and mid-infrared (NIR \& MIR respectively).

The \emph{Wide-field Infrared Survey Explorer} \citep[\wise, ][]{2010AJ....140.1868W, 2012wise.rept....1C} is a space-based observatory that has imaged the entire sky in the MIR (3.4, 4.6, 12, and 22$\mu$m). Additionally, \wise\, has been positionally matched to the Two-Micron All-Sky Survey (\twomass), a four-year mission characterizing the full sky in the NIR. Thus, the \wise\, catalog presents with hundreds of millions of sources with photometry of unprecedented sensitivity in the NIR and MIR--ideal for capturing \agb\, stars at Galactic distances. In Section~\ref{sec:data}, we describe in detail the data that we use for our study.

In Section~\ref{sec:criteria}, we describe the color-color criteria used to isolate \agb\, stars in the \wise\, dataset, and the color-magnitude relationships that were derived for these stars from the Large Magellanic Cloud and the \mw\, bulge.
In Section~\ref{sec:distribution}, we describe the spatial density distribution of \agb\, candidates in the Milky Way.
Our conclusions can be found in section~\ref{sec:conclusions}.

>>>>

\section{Introduction}
%The formation of galaxies like the \mw\, was long thought to be a steady process that created a smooth distribution of stars. This standard view was exemplified by the \cite{1980ApJS...44...73B} and \cite{1989ARA&A..27..555G} models, described in detail by, e.g., \cite{1993ARA&A..31..575M}. It was further motivated by observations of other galaxies, as well as what little information was available from \emph{High Precision Parallax Collecting Satellite} \citep[\emph{HIPPARCOS}, ][]{1984SSRv...39....1K} and smaller pencil-beam surveys. In these, 
The structure of the \mw\, has for many years been uncertain, with historical models assuming three discrete components described by simple analytic expressions: the thin disk, thick disk, and halo \citep{1980ApJS...44...73B, 1989ARA&A..27..555G,1993ARA&A..31..575M}. 
The advent of the \emph{Sloan Digital Sky Survey} \citep[\sdss, ][]{2000AJ....120.1579Y} has since provided more detail, using accurate digital multi-band optical photometry across a quarter of the sky, as well as the largest optical spectroscopic catalog thus far known. 
%This new influx of data enabled the development and application of photometric parallax methods, using color-magnitude relations to estimate stellar distances.
These new data led to the large scale ``tomography" of the \mw\; via stellar distributions in the 7-dimensional space spanned by spatial coordinates \citep{2008ApJ...673..864J}, velocity components \citep{2010ApJ...716....1B}, and metallicity \citep{2008ApJ...684..287I}. The resulting maps revealed rich, complex substructure in the distribution of the \mw's stars \citep[e.g.][]{2000AJ....120..963I,2000ApJ...540..825Y,2001ApJ...554L..33V,2002ApJ...569..245N,2003ApJ...599.1082M,2006ApJ...642L.137B,2006ApJ...651L..29G,2006AJ....132..714V}, deeply shaking the existing view of the Galaxy.

In order to move forward from where \sdss\, tomography left off, we require observations that span an area larger than that of \sdss\,, with Galactic objects that can be seen through interstellar dust out to large distances. Stars from the Asymptotic Giant Branch (\agb) are perfect candidates to serve as these probes to the \mw. \agb\, stars represent the last stage of evolution for stars between 0.8 and 8 $M_\odot$ \citep{1983ARA&A..21..271I, 2005ARA&A..43..435H} -- the mass-range with the highest number of stars as inferred from the  \cite{2001MNRAS.322..231K} initial mass function that can also reach the final stages of stellar evolution within a galactic timescale. Because of this, they are bound to reside throughout the galaxy wherever other stars are present. This phase of evolution is marked by two distinct periods: the early \agb\, (E-\agb)  and the thermally-pulsing \agb\, phases. During the latter of the two, \agb\, stars produce substantial dust-driven stellar winds \citep[$10^{-7} < \dot{M} < 10^{-4}$ $M_\odot$ yr$^{-1}$,][]{2002A&A...391.1053O} rich in oxides (SiO, Al$_2$O$_3$, etc.) and carbon-rich molecules (SiC, AmC, etc.), with the chemical dominance being highly dependent upon the metallicity of the host galaxy \citep{2005A&A...434..691M}. Galaxies such as the Milky Way are expected to have a substantial population of oxygen-rich AGB stars \citep{1985A&A...152L...1H}, whereas low-metallicity galaxies such as the Magellanic clouds have been shown to possess AGB populations dominated by carbon-rich stars \citep{2011AJ....142..103B}. In both cases, the other species of AGB star is rarely seen, as richness in one chemical type (e.g. oxides)  necessitates the almost complete capture of the other chemical type (e.g. carbon) in CO \citep{1983ARA&A..21..271I}. Over time, these chemically-rich winds create vast circumstellar shells that, when warmed by the stellar photosphere, emit in the near- and mid-infrared (NIR \& MIR respectively). Although the molecular species present in oxygen- and carbon-rich winds are vastly different, emission from both types produce strong IR emission near 10$\mu$m, visible out to the Magellanic clouds \citep{2011A&A...534A..79I} and beyond. Thus, if these stars can be pinpointed by a survey with a wide area of observation and high positional accuracy, they can be used as markers for a map of the Milky Way \citep{2013RAA....13..323T}.

Such a survey can be found in the \emph{Wide-field Infrared Survey Explorer} \citep[\wise, ][]{2010AJ....140.1868W, 2012wise.rept....1C}, a space-based observatory that has imaged the entire sky in the MIR (3.4, 4.6, 12, and 22$\mu$m). Additionally, \wise\, has been positionally matched to the Two-Micron All-Sky Survey (\twomass), a four-year mission characterizing the full sky in the NIR. Thus, the \wise\, catalog presents with hundreds of millions of sources with photometry of unprecedented sensitivity and positional accuracy in the NIR and MIR--ideal for capturing \agb\, stars at Galactic distances. In this paper, we use samples of known Galactic and Magellanic \agb\, stars to formulate color-color criteria with \wise\, and \twomass\, photometry that can produce a reliable catalog of IR \agb\, candidates. We then use these candidates in conjunction with estimates of Galactic dust extinction along the line of sight to produce a Galaxy-wide number density distribution of \agb\, stars.

In Section~\ref{sec:data}, we describe in detail the data that we use for our study. In Section~\ref{sec:criteria}, we describe the color-color criteria used to isolate \agb\, stars in the \wise\, dataset, and the color-magnitude relationships that were derived for these stars from the Large Magellanic Cloud and the \mw\, bulge.
In Section~\ref{sec:distribution}, we describe the spatial density distribution of \agb\, candidates in the Milky Way.
Our conclusions can be found in section~\ref{sec:conclusions}.
